Type-Safe Client-Server Communication in Web Applications

### 2014-02-27 paper.tex
Add blurb about our solution being client centric.
Introduce some subsections.

Several attempts have been made at tackling this problem, but none is, in our
view, satisfactory. In this paper, we propose a programming model in which a
web application is written as a single program from which client and server
binaries are generated during compilation. Type annotations in the source
program control which parts are executed on the server and which are executed
on the client, and the two communicate using type safe RPC calls. Functions
which are not explicitly declared as server-side or client-side are usable by
either side. Our solution is implemented as a Haskell library requiring minimal
to no compiler support, ensuring portability and modularity.

@@

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support.
\end{itemize}

\paragraph{A first example} While explaining of the properties of our solution
is all well and good, nothing compares to a good old Hello World example to
convey the idea. We begin by implementing a function which prints a greeting
to the server's console.

@@

\paragraph{Choosing a compiler} Haste is by no means the only
Javascript-targeting compiler for a purely functional language. In particular,
the GHC-based GHCJS\ \cite{ghcjs} and UHC\ \cite{uhc} compilers are both
capable of compiling standard Haskell into Javascript; the Fay\ \cite{fay}
language was designed from the ground up to target the web space using a subset
of Haskell; and there exist solutions for compiling Erlang\ \cite{jserlang} and
Clean\ \cite{jsclean} to Javascript as well.

@@

\paragraph{Design of the Haste compiler} Haste is first and foremost intended
to be a practical compiler; the ability to produce code which is lean and fast
enough to be used in the real world is critical. To this end, we make several
design choices and tradeoffs, some of which would be viewed as detrimental in a
more purist context.

@@

\paragraph{Implementation} As previously mentioned, Haste offloads much of the
heavy lifting of compilation - parsing, type checking, intermediate code
generation and many optimizations - onto GHC, and takes over code generation
after the STG generation step, at the very end of the compilation process.
STG\ \cite{stg} is the last intermediate representation used by GHC before
the final code generation takes place and has several benefits for use as
Haste's target language:

@@

@@

\paragraph{Interfacing with Javascript} While Haste supports the Foreign
Function Interface inherited from GHC, with its usual features and limitations
\ \cite{ffi}, it is often impractical to work within the confines of an
interface designed for communication on a very low level. For this reason
Haste sports its own method for interacting with Javascript as well, which
allows the programmer to pass any value back and forth between Haskell and
Javascript, as long as she can come up with a way to translate this value
between its Haskell and Javascript representations. Of course, not performing
any translation at all is also a valid ``translation'', which allows Haskell
code to store any Javascript value for later retrieval without inspecting it
and vice versa. The following example implements mutable variables using this
custom Javascript interface.

\begin{code}
import Haste.Foreign

@@

getMutable :: Marshal a => MutableVar a -> IO a
getMutable = ffi "(function(m) {return m.val;})"
\end{code}

@@

@@

\paragraph{Limitations} Unlike any of the related works, our approach takes a
firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! would be relatively
straightforward.

While our solution completely safe from the invocation point of the
\lstinline!runApp! function onward, capturing IO-dependent variables introduced
before this invocation may lead to the client and server execution paths
diverging. Consider the following example:

@@

>>>>

Several attempts have been made at tackling this problem, but none is, in our
view, satisfactory. In this paper, we propose a functional programming model in
which a web application is written as a single program from which client and
server binaries are generated during compilation. Type annotations in the source
program control which parts are executed on the server and which are executed
on the client, and the two communicate using type safe RPC calls. Functions
which are not explicitly declared as server-side or client-side are usable by
either side. Our solution is implemented as a Haskell library requiring minimal
to no compiler support, ensuring portability and modularity.

@@

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support.
\end{itemize}

\subsection{A first example}

While explaining of the properties of our solution is all well and good,
nothing compares to a good old Hello World example to convey the idea.
We begin by implementing a function which prints a greeting to the server's
console.

@@

\subsection{Choosing a compiler}

Haste is by no means the only Javascript-targeting compiler for a purely
functional language. In particular, the GHC-based GHCJS\ \cite{ghcjs} and
UHC\ \cite{uhc} compilers are both capable of compiling standard Haskell into
Javascript; the Fay\ \cite{fay} language was designed from the ground up to
target the web space using a subset of Haskell; and there exist solutions for
compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to Javascript as
well.

@@

\subsection{Design of the Haste compiler}

Haste is first and foremost intended to be a practical compiler; the ability
to produce code which is lean and fast enough to be used in the real world is
critical. To this end, we make several design choices and tradeoffs, some of
which would be viewed as detrimental in a more purist context.

@@

\subsection{Implementation overview}

As previously mentioned, Haste offloads much of the heavy lifting of
compilation - parsing, type checking, intermediate code generation and many
optimizations - onto GHC, and takes over code generation after the STG
generation step, at the very end of the compilation process. STG\ \cite{stg} is
the last intermediate representation used by GHC before the final code
generation takes place and has several benefits for use as Haste's target
language:

@@

\subsection{Data representation}

@@

\subsection{Interfacing with Javascript}

While Haste supports the Foreign Function Interface inherited from GHC, with
its usual features and limitations \ \cite{ffi}, it is often impractical to
work within the confines of an interface designed for communication on a very
low level. For this reason Haste sports its own method for interacting with
Javascript as well, which allows the programmer to pass any value back and
forth between Haskell and Javascript, as long as she can come up with a way to
translate this value between its Haskell and Javascript representations.
Of course, not performing any translation at all is also a valid
``translation'', which allows Haskell code to store any Javascript value for
later retrieval without inspecting it and vice versa. The example given in
figure \ref{lst:ffi} implements mutable variables using this custom Javascript
interface.

\begin{listingfloat}
\begin{code}
import Haste.Foreign

@@

getMutable :: Marshal a => MutableVar a -> IO a
getMutable = ffi "(function(m) {return m.val;})"
\end{code}
\caption{Mutable variables with \lstinline!Haste.Foreign!}
\label{lst:ffi}
\end{listingfloat}

@@

\subsection{Related work}

@@

\subsection{Limitations}

\paragraph{Client-centricity} Unlike any of the related works, our approach
takes a firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client-side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! would be relatively
straightforward.

\paragraph{Preventing divergence} While our solution completely safe from the
invocation point of the \lstinline!runApp! function onward, capturing
IO-dependent variables introduced before this invocation may lead to the client
and server execution paths diverging. Consider the following example:

@@

\paragraph{Environment consistency} As our programming model uses two different
compilers to generate client and server code, it is crucial to keep the package
environments of the two in sync. A situation where, for instance, a module is
visible to one compiler but not to the other will render many programs
uncompilable until this inconsistency is fixed.

This kind of divergence can be worked around using conditional compilation, but
is highly problematic even so; using a unified package database between the two
compilers, while problematic due to the differing natures of native and
Javascript compilation respectively, would be a significant improvement in this
area.
